---
title: "APC Analysis"
author: "shelloren"
date: "Last updated: `r format(Sys.Date(), '%B %e, %Y')`"
output: 
  html_document:
    toc: true
    toc_depth: 2
    toc_float: 
        collapsed: false
    number_sections: true
    theme: flatly
    highlight: tango
    css: style.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(scipen = 9999)
rm(list = ls())
```

# Background

## Disclaimer

The following coursebook is produced by the team at [Algoritma](https://algorit.ma) for its Data Science Academy internal training. The coursebook is intended for a restricted audience only, i.e. the individuals and organizations having received this coursebook directly from the training organization. It may not be reproduced, distributed, translated or adapted in any form outside these individuals and organizations without permission. 

## Libraries and Setup

You will need instal use few package to do APC analysis if it's not already downloaded onto your machine.

```{r lib, message=FALSE}
library(broom)
library(tidyverse)
```

## Training Objectives

The primary objective of this course is to provide a comprehensive introduction to the science of APC analysis. The syllabus covers:

- **APC Analysis**
- Workflow   
- APC in R  

# APC Analysis

Age-period-cohort (APC) analysis is widely used for capturing the characteristics in time-varying elements, and mostly used in epidemiology and demographic fields. In particular, APC analysis discerns three types of time varying phenomena[^1]: 

* **Age effects**:
    Age effect is variations linked to biological and social processes of aging specific to individuals. They include physiologic changes and accumulation of social experiences linked to aging, but unrelated to the time period or birth cohort to which an individual belongs. 
    
* **Period effects**:
    Period effect is the result from external factors that equally affect all age groups at a particular calendar time. It could arise from a range of environmental, social and economic factors e.g. war, famine, economic crisis. Methodological changes in outcome definitions, classifications, or method of data collection could also lead to period effects in data.

* **Cohort effects**:
    Cohort effect is variations resulting from the unique experience/exposure of a group of subjects (cohort) as they move across time.  In epidemiology, a cohort effect is conceptualized as an interaction or effect modification due to a period effect that is differentially experienced through age-specific exposure or susceptibility to that event or cause.

In contrast to this conceptualization of cohort effect as an effect modification in epidemiology, sociological literature consider cohort effect as a structural factor representing the sum of all unique exposures experienced by the cohort from birth. In this case, age and period effect are conceived as confounders of cohort effect and APC analysis aims to disentangle the independent effect of age, period and cohort. Most of the APC analysis strategies are based on the sociological model of cohort effect, conceptualize independent effect of age, period and cohort effect.

**Identification problem in APC:** APC analysis aims at describing and estimating the independent effect of age, period and cohort on the outcome under study. The different strategies used aims to partition variance into the unique components attributable to age, period, and cohort effects. However, there is a major impediment to independently estimating age, period, and cohort effects by modeling the data which is know as the “identification problem” in APC. This is due to the exact linear dependency among age, period, and cohort: Period – Age = Cohort; that is, given the calendar year and age, one can determine the cohort (birth year). The presence of perfectly collinear predictors (age, period and cohort) in a regression model will produce a singular non-identifiable design matrix, from which it is statistically impossible to estimate unique estimates for the three effects.


## Guideline for estimating APC models[^2]:

1. Descriptive data analysis by graphical representation of data is the first step in an APC analysis. This helps in qualitative assessment of patterns of time-based variations

2. Rule out that data can be explained by any single factor or two factor model of age, time period and cohort. A Goodness-of-fit statistics is often used to compare between reduced log linear models: three separate models for age, period and cohort effects; and three two-factor models, one for each of three possible pairs of effects, namely, AP, AC, and PC effects models. All of these models then are compared with a full APC model where all three factors are simultaneously controlled.Two most commonly used penalized-likelihood model selection criteria namely, the Akaike information criterion (AIC) and the Bayesian information criterion (BIC), are used to evaluate the model since likelihood ratio tests tend to favor models with a larger number of parameters. BIC and AIC both adjust the impact of model dimensions on model deviances.

3. If the descriptive analyses indicate that all three of the A, P, and C dimensions is not operative, then the analysis can be completed with a reduced model that omits the non-operative dimension and there is no identification problem.

4. If, however, these analyses suggest that all three dimensions are at work use one of the specific methods of APC analysis


## APC 

I've prepared the following data originally made available in [kaggle](https://www.kaggle.com/russellyates88/suicide-rates-overview-1985-to-2016). We'll read the data into our workspace:

```{r, message=FALSE}
# load data
master <- read_csv("data/master.csv")

# quick check
glimpse(master)
```

The data had 27.820 obs and 12 Var, here some description to get insight of the data:  

- `sucides_no` : number of suicedes each year.
- `population` : country-year-sex-age level.
- `suicides/100k pop` :
- `country-year` :
- `HDI for year` :
- `gdp_for_year ($)` :
- `gdp_for_capita ($)` :
- `generation` : based on age grouping average.

We need to make APC variable to start with:

```{r}
# make apc variable
fix <- master %>%
  mutate(
    age = str_replace_all(age, " years", "") %>%
      str_replace_all("\\+", "\\-75") # seperate "years" and "+" from age
  )  %>% 
  separate(age, into = c("age_lower","age_upper"), sep = "-") %>% # divide age range
  mutate_at(vars(matches("age")), funs(as.numeric(.))) %>% # change to numeric
  mutate(
    age = round((age_lower + age_upper) / 2),
    period = year,
    cohort = period - age
  ) %>% # create APC
  mutate_at(vars(age, period, cohort), funs(as.factor(.))) %>% # turn it to factor
  select(suicides_no, age, period, cohort, sex, country) 

knitr::kable(fix) %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed"))

```

Now we have our data that contained APC variables. For the next step

```{r}
apc_lm <- lm(suicides_no ~ age + period + cohort, fix)
apc_female_lm <- lm(suicides_no ~ age + period + cohort, apc_female)
apc_male_lm <- lm(suicides_no ~ age + period + cohort, apc_male)

```

```{r}
apc_tidy <- tidy(apc_lm)
apc_female_tidy <- tidy(apc_female_lm)
apc_male_tidy <- tidy(apc_male_lm)
```

```{r}
apc_tidy %>%
  slice(-1) %>% 
  mutate(
    key = str_replace_all(term, "[:digit:]", ""),
    value = str_replace_all(term, "[:alpha:]", "") %>% as.numeric()
  ) %>% 
  ggplot(aes(value, estimate)) +
    geom_line() +
    facet_wrap(~ key, scales = "free", ncol = 1) + 
    labs(title = "Age-Period-Cohort Analysis")

```

# Annotations
[^1]: Yang Y, Schulhofer‐Wohl S, Fu WJ, Land KC. The Intrinsic Estimator for Age‐Period‐Cohort Analysis: What It Is and How to Use It1. American Journal of Sociology 2008;113(6):1697-736.
[^2]: Yang, Yang, and Kenneth C. Land. Age-period-cohort analysis: new models, methods, and empirical applications. CRC Press, 2013